Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2014-09-24 draft_lm1.tex
updated the introduction and references

Multiple factors contribute to the dificulty in estimating the number of 
susceptibles in the population, first and foremost is the fact that dengue 
confers permanent immunity within serotypes but with only short-lived 
cross-immunity.
Secondly, there are a larger number of asymptomatic cases in 
any epidemic, which nevertheless acquire immunity.
Lastly Demography and 
migrations affect the number of susceptible in ways which are not easy to 
determine.

@@

In this paper, we estimate the initial number of susceptibles right before 
every epidemic in the last 18 years.
For this we use a simplified model of 
dengue transmission based on a Susceptible-infecctious-Removed (SIR) model.
We then fit the 
model output to observed data to estimate the $S_0$ before every 
recorded epidemic in the Brazilian city of Rio de Janeiro. Rio de Janeiro has 
bee through many Dengue epidemics of various serotypes and sizes.

>>>>

Multiple factors contribute to the dificulty in estimating the number of 
susceptibles in the population, first and foremost is the fact that dengue

confers permanente immunity within serotypes but with only short-lived 
cross-immunity. Secondly, there are a larger number of asymptomatic cases in 
any epidemic, which nevertheless acquire immunity, Thirdly, under-reporting is 
a serious issue in Brazil. Duarte and Franca\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting in Belo-Horizonte, Brazil to be of 63\%, 
meaning that aproximately 37\% of the cases go unreported. Lastly, demography 
and migrations affect the number of susceptible in ways which are not easy to
determine.

@@

In this paper, we estimate the number of susceptibles, allowing for 
discontinuities right before every epidemic in the last 18 years. THis way we 
can account for replacement of the predominant serotype, which can redefine 
susceptibility for the following season.

To accomplish this we use a simplified model of 
dengue transmission based on a Susceptible-infecctious-Removed (SIR) model. 
Since Dengue is not caused by a single pathogen, but by four separate strains 
with limited cross-immunity, when we talk about number of susceptibles, in the 
context of a single serotype SIR model, we are actually referring to the sum of 
the susceptibles to the currently circulating serotypes at any given time.

Some information is available about the predominant serotypes during the 
period of study\cite{macedo_virological_2013}.

The estimation of the  number of susceptibles have been attempted 
before, for other diseases\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}. These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.\todo[inline]{anyone else?} In the case of Dengue the 
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergenge of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation of the dynamics. 
Adopting a multi-strain transmission model, does not help because of 
insufficient information about the proportions of each serotype in case data.

The authors have applied similar methods as used here, to estimate the number 
of susceptibles for the predominant strain of influenza at the beginning of  
influenza seasons in europe\cite{pone2011}.
